%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% (C) Copyright 1996, 2020-2021 Kenneth Geisshirt <geisshirt@gmail.com>
%%
%% Licensed under Creative Commons Attribution-ShareAlike 4.0 International.
%% You may obtain a copy of the license at
%% https://creativecommons.org/licenses/by-sa/4.0/
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction and basic definitions}
\label{chap:intro}

Let us begin by defining and limiting our term "chemical kinetics". As already mentioned in the preface, chemical kinetics is not about equilibrium, but the dynamics of chemical reactions.

We can choose two distinct approaches towards chemical kinetics, namely the microscopic or the macroscopic one. The difference is the size of the system studied. In the microscopic approach we will study one molecule reacting with another molecule, while the macroscopic approach is a more phenomenological one. Here we study many molecules, maybe 1 mole (remember that 1 mole is $6.022\times 10^{23}$ molecules).

The present text approach chemical kinetics from the macroscopic viewpoint.

The dynamics of the macroscopic viewpoint is the change in concentration. In other words, we should imaging that we are monitoring the concentration of a given species as a function of time. For example imaging the reaction $\species{I_2} + \species{I} \rightarrow \species{I_3}$. In this case we could measure the concentration of triodide \ie $[\species{I_3}](t)$ where the $t$ indicates that the concentration is measured as a function of time.

I have to come with a warning before we begin our journey into the land of chemical kinetics. Often we will not use real chemical species, but only abstract ones like $A$ and $B$. The convention is that capital letters are always chemical species. In the beginning this may seem strange, but once learned, we can apply our results to many (real) reactions.

\section{Velocity of a reaction}

In order to do a more formalized (\ie mathematical) analysis of chemical reactions, it is not enough just to measure the concentration of the species involved. A more convenient approach is to begin the analysis by discussing the velocity of the reaction.

The most general form we can write up a chemical reaction as

\[
  aA + bB + \cdots + xX + yY + \cdots \rightarrow 0
\]

where $A$, $B$ are reactants, $a$, $b$ their stoichiometric coefficients, $X$, $Y$ the products, and $x$, $y$ their stoichiometric coefficients. Remember, our convention is that the stoichiometric coefficients for reactants are negative, and positive for products.

We will \textit{define} the velocity, $v$ of this reaction as

\begin{equation}
  \label{eq:velocity}
  v = \frac{1}{a}\dt{[A]} = \frac{1}{b}\dt{[B]} = \frac{1}{x}\dt{[X]} = \frac{1}{y}\dt{[Y]}
\end{equation}

\begin{example}
  We will illustrate the concept of velocity by looking at the decomposition of hydrogen peroxide \ie

  \[
      2\species{H_2O_2} \rightarrow 2\species{H_2O} + \species{O_2}
  \]

  We have three species ($\species{H_20_2}$, $\species{H_2O}$, and $\species{O_2}$), and the stoichiometric coefficients are —2, 2 and 1, respectively. The velocity is therefore

  \[
    v = -\frac{1}{2}\dt{[\species{H_2O_2}]} = \frac{1}{2}\dt{[\species{H_2O}]} = \dt{[\species{O_2}]}
  \]
\end{example}

So far we have not said anything about the velocity. If we are able to give an expression for $v$, we would have a differential equation\footnote{To be more precise: an ordinary differential equation of first order}. By solving this differential equation we would find the concentration as function of time (which is comparable to experiments). It is important to stress that the expression for the velocity is an experimental fact.

Often we call the expression for the velocity of the reaction for the \textit{rate law}.

\section{Elementary reactions}
\label{sec:elementaryReactions}

All chemical reactions can be divided into two classes, elementary and non-elementary reactions. The elementary reactions are the ones that happen in Nature (usually our laboratory). Any non-elementary reaction is the sum of a number of elementary reactions.

Elementary reactions usually only have one or two reactants. But as we will see later the reaction $\species{H_2} + \species{Br_2} \rightarrow 2\species{HBr}$ is not an elementary reaction as one might first think. Enzyme reactions are never written down in terms of elementary reactions while gas phase reactions usually are.

Elementary reactions have one nice feature; we can - almost automatically - write down an expression for the velocity of the reaction. Or we could say it the opposite way: if the reaction follows a certain expression for the velocity (an expression we will discuss below), we know it is an elementary reaction.

Let us use a generalized reaction \ie

\[
  aA + bB + \cdots \rightarrow xX + yY \cdots
\]

where the stoichiometric coefficients for both the reactants and the products are positive!

For an elementary reaction the velocity is

\begin{equation}
  \label{eq:elementaryReaction}
  v = k[A]^a[B]^b\cdots
\end{equation}

The constant $k$ is a proportionality constant which we call the \textit{rate constant}.

\begin{example}
  We continue our example from previous section \ie the decomposition of hydrogen peroxide:

  \begin{equation}
    \label{eq:decompH2O2}
    2\species{H_2O_2} \rightarrow 2\species{H_2O} + \species{O_2}
  \end{equation}

  There is only one reactant, $\species{H_2O_2}$, and the velocity is

  \begin{equation}
    \label{eq:decompH2O2velocity}
    v = k\cdot [\species{H_2O_2}]^2
  \end{equation}
\end{example}

In the example we can see one of the bad habits used in chemical kinetics. The concentration of hydrogen peroxide is written as $[\species{H_2O_2}]$, but this notation does not emphasize that the concentration charges with time. The correct way of writing the concentration of hydrogen peroxide is $[\species{H_2O_2}](t)$, but we will rarely do that.

Some chemical kineticists call reactions where the velocity is written as in equation \ref{eq:elementaryReaction} kinetics consistent with the \textit{law of mass action}. The reason for this is that the expression for the velocity reminds us of the one used in the treatment of chemical equilibrium. Of course nobody uses so long a term everyday -- one just says: mass action kinetics.

Ending the discussion of the concept of elementary reactions we will introduce the order of a reaction. The term "order" is only meaningful when we are talking about mass action kinetics. The order of a reaction is defined as the sum of the stoichiometric coefficients of the reactants (or more precisely, the sum of the exponents in the rate law). In the example of the decomposition of hydrogen peroxide, the order is 2.